ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-29 doc/paper.tex
split lit review

\graphicspath{{images/}}

@@

@@

@@

The code in this paper was used as a reference for the phase vocoder implementation in ScPy.
Additionally, some of the extra techniques mentioned could improve the quality or flexibility of
the implementation in the future.

Keiler and Sylvain~\cite{keiler2002survey} present a survey of 6 techniques for accurate analysis
of stationary sounds. The first method, plain FFT, uses a Hann window and assumes the frequencies
of the sinusoids at centered within each bin. Parabolic interpolation notices that the shape of a
magnitude peak looks like a parabola, and thus attempts to fit a parabola to the magnitudes of bins
nearby each local maxima, yielding more accurate results than a plain FFT\@. The triangle algorithm
is similar to the parabolic interpolation method but fits a triangular function instead of a
parabola. Spectral reassignment uses the derivative of the analysis window to estimate the true
values of the bins as the center of gravity of the energy of each bin. The derivative algorithm
also uses the derivative of the analysis window to determine both a more accurate frequency and
amplitude. It can be further extended to use the first nth derivatives. The phase vocoder uses the
phase information of two successive FFTs to improve the frequency resolution by calculating the
phase derivative. Comparison of the techniques found that the spectral reassignment, derivative
algorithm, and phase vocoder methods had the best performance. Surprisingly, those three techniques
produced almost identical results, suggesting that they may have equivalent theoretical basis.

The two other techniques that performed similarly to the phase vocoder will be interesting to
analyze in more detail since they may present a solution to the \textit{phasiness} problem of the
phase vocoder due their lack of persistent state. However, they require extra processing so it is
not clear if the trade-off will be worthwhile.

Audio morphing is defined as a gradual change from one sound to the other. While morphing can be
achieved to some extent simply by cross fading two signals, perceptually this sounds simply like
one sound is becoming quieter as the other is becoming louder. Sethares and Bucklew
\cite{sethares2012kernel} propose density and kernel based techniques that aim to provide a more
convincing morph operation where there is no clear separation between the two sounds at any point
during the morphing process. The method allows the use of different kernel functions to provide
different kinds of morphing. These include \textit{Cross-fade}, \textit{Heat equation}, and
\textit{Harmonic integrity}. The morphing technique is implemented using a phase vocoder. These
techniques could be implemented in Python as a future extension to ScPy.

Laroche and Dolson~\cite{laroche1999improved} present an improvement to the phase vocoder which
aims to prevent the ``phasiness'' problem where different parts of a sound lose their phase
coherence, as well as provide a significant decrease in computational cost. The paper recognizes
that there are two types of phase coherence that must be preserved in any STFT based algorithm to
ensure that the reproduced sound is without artifacts: horizontal and vertical. The basic phase
vocoder algorithm is built to preserve horizontal phase coherence --- but it does so at the expense
of vertical phase coherence. To remedy this, they build on prior work by presenting two new
phase-locking techniques.

@@

The phase vocoder currently implemented in ScPy uses the standard algorithm. Future experiments
could implement many of the proposed modifications to the phase vocoder to improve sound quality.

Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved (at a high computational cost). Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

Using Theano, the sliding DFT and sliding phase vocoder could be implemented easily in ScPy. It
remains unknown if it will be possible to run them in real time, but even if not this future
extension could add the possibility to experiment with many novel effects.

Klingbeil~\cite{klingbeil2005software} describes the techniques used in a high quality graphical
spectral editing and resynthesis program called SPEAR\@. The FFT analysis is done taking into
account the fundamental frequency of the sound to analyze, so that the transform can be optimized
to accurately capture as many partials as possible. Spectral peaks are then selected and very quiet
peaks are discarded. A linear prediction method is used to follow the peaks over time. The data
structure for storage of spectral data involves division of time into a number of frames which keep
track of which partials are active during that time, so that the search for a specific partial can
be narrowed. User interface design choices are also covered in detail.

Although the interface and granularity of SPEAR differs significantly from those in ScPy, the
techniques for high quality resynthesis presented should prove useful for improving the quality of
the sound output.

While the Fourier transform is by far the most commonly used technique for spectral manipulation in
digital audio effects, there exist numerous other transformations which may be useful as well to
achieve certain audio effects. One such technique is the Mellin transform, described in a audio
effects context by De Sena and Rocchesso~\cite{de2004fast}. The Mellin transform has a scale
invariance property that is analogous to the shift invariance property of the Fourier transform. It
can be computed quickly by applying two extra steps before doing a Fourier transform: exponential
time warping and exponential multiplication. A few simple audio effects are presented using the
Mellin transform, time stretching, filtering and random phase deviation (whisperization).

As the Mellin transform can be implemented simply as an extension to the Fourier transform, it
should be fairly easy to add support for it in ScPy, and it will hopefully enable the use of
some different audio effects than can be achieved easily with the Fourier transform. However, the
paper mentions that a short-time Mellin transform has not been attempted, so some further work may
need to be done in that area first.

Collins et al.~\cite{collins2003live} present an introduction to the field of live coding, where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the just in time library.

@@

@@

The ideas in Nilson's paper shed light on some of the challenges faced in this project.
One such challenge is the amount of writing required to produce an interesting result from a blank
slate. We intend to minimise this by providing many effect primitives and composition
mechanisms. However, it will be important to ensure the language is accessible to ensure it can be
quickly learned to a level of expertise where lack of technical knowledge is not an impediment.

Domain-Specific Languages (DSLs) are small programming languages designed for use in a specific
problem domain. They often lack general purpose programming constructs and have a syntax which is
very concise for describing programs in their domain. Van Deursen et al.~\cite{van2000domain}
present a survey of available DSL literature, along with the pros and cons of using a DSL, example
DSLs, design methodology and implementation approaches. The biggest advantage of DSLs is that they
enable expression of solutions in a way that is common for the problem domain. However, they can be
costly to develop and difficult to design. Implementation techniques include the classical approach
of interpretation or compilation, embedded languages / domain-specific libraries, preprocessing, or
compiler extension. Additionally, the references of the paper each include a short summary.

This review reveals many of the possible options for the DSL used within the project. Although we
decided not to design a DSL for our project, analyzing the design and implementation techniques
presented here will be helpful for future work in this area.

As Python is an easily extendible general purpose language, it has applications in many scientific
fields. Glover et al.~\cite{glover2011python} present a discussion of the use of Python for audio
signal processing. They note that while audio processing libraries exist for many languages,
scripting languages such as Python or MATLAB see widespread adoption because of the speed at which
prototypes can be built. General purpose Python libraries such as NumPy, SciPy and Matplotlib
include many essential functions for audio processing, such as file IO, Fourier transforms, signal
processing and interpolation. There are additionally more specific libraries such as Simpl, a
sinusoidal modelling library, and Modal, an onset detection library. Any of these libraries can
easily be used within ScPy simply by installing and importing them.

Another open-source Python library which could be used in ScPy is librosa~\cite{mcfee2015librosa}
by McFee et al.\ librosa is a music signal analysis library designed to help transition music
information retrieval (MIR) researchers into using Python, as well as make MIR techniques readily
available to all Python programmers.\ librosa's core design choices and conventions are documented
in detail to inspire future library authors to follow similarly good coding practice. As well as
standard audio analysis functions, librosa also includes support for effects and audio IO.\@

librosa has quite an extensive list of features and would likely cover all of the audio related
functionality which is missing from NumPy. It can be used with ScPy simply by installing and importing
it.

The ixi lang~\cite{magnusson2011ixi} is a live coding language. It is implemented as an interpreter
built in SuperCollider, a choice which allows for SuperCollider and ixi lang code to be mixed
together in a similar manner to ScPy. The ixi lang is built to reduce the startup time in a live
coding performance and enable faster musical experimentation through syntax more concise than
SuperCollider. The paper also contains user feedback on the usability of the ixi lang.

@@

The choice of SuperCollider was largely because it is one of the most widely used audio programming
environments. Using SuperCollider also enabled the reuse of many parts of its codebase. Since it
already has STFT based operations, it was a good target for extension purposes in that area.

>>>>

@@

\subsection{NIME}

@@

\subsection{Spectral audio processing}

@@

The code in this paper was used as a reference for the phase vocoder implementation in ScPy.
Additionally, some of the extra techniques mentioned could improve the quality or flexibility of
the implementation in the future.
\\\par
Keiler and Sylvain~\cite{keiler2002survey} present a survey of 6 techniques for accurate analysis
of stationary sounds. The first method, plain FFT, uses a Hann window and assumes the frequencies
of the sinusoids at centered within each bin. Parabolic interpolation notices that the shape of a
magnitude peak looks like a parabola, and thus attempts to fit a parabola to the magnitudes of bins
nearby each local maxima, yielding more accurate results than a plain FFT\@. The triangle algorithm
is similar to the parabolic interpolation method but fits a triangular function instead of a
parabola. Spectral reassignment uses the derivative of the analysis window to estimate the true
values of the bins as the center of gravity of the energy of each bin. The derivative algorithm
also uses the derivative of the analysis window to determine both a more accurate frequency and
amplitude. It can be further extended to use the first nth derivatives. The phase vocoder uses the
phase information of two successive FFTs to improve the frequency resolution by calculating the
phase derivative. Comparison of the techniques found that the spectral reassignment, derivative
algorithm, and phase vocoder methods had the best performance. Surprisingly, those three techniques
produced almost identical results, suggesting that they may have equivalent theoretical basis.

The two other techniques that performed similarly to the phase vocoder will be interesting to
analyze in more detail since they may present a solution to the \textit{phasiness} problem of the
phase vocoder due their lack of persistent state. However, they require extra processing so it is
not clear if the trade-off will be worthwhile.
\\\par
Audio morphing is defined as a gradual change from one sound to the other. While morphing can be
achieved to some extent simply by cross fading two signals, perceptually this sounds simply like
one sound is becoming quieter as the other is becoming louder. Sethares and Bucklew
\cite{sethares2012kernel} propose density and kernel based techniques that aim to provide a more
convincing morph operation where there is no clear separation between the two sounds at any point
during the morphing process. The method allows the use of different kernel functions to provide
different kinds of morphing. These include \textit{Cross-fade}, \textit{Heat equation}, and
\textit{Harmonic integrity}. The morphing technique is implemented using a phase vocoder. These
techniques could be implemented in Python as a future extension to ScPy.
\\\par
Laroche and Dolson~\cite{laroche1999improved} present an improvement to the phase vocoder which
aims to prevent the ``phasiness'' problem where different parts of a sound lose their phase
coherence, as well as provide a significant decrease in computational cost. The paper recognizes
that there are two types of phase coherence that must be preserved in any STFT based algorithm to
ensure that the reproduced sound is without artifacts: horizontal and vertical. The basic phase
vocoder algorithm is built to preserve horizontal phase coherence --- but it does so at the expense
of vertical phase coherence. To remedy this, they build on prior work by presenting two new
phase-locking techniques.

@@

The phase vocoder currently implemented in ScPy uses the standard algorithm. Future experiments
could implement many of the proposed modifications to the phase vocoder to improve sound quality.
\\\par
Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved (at a high computational cost). Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

Using Theano, the sliding DFT and sliding phase vocoder could be implemented easily in ScPy. It
remains unknown if it will be possible to run them in real time, but even if not this future
extension could add the possibility to experiment with many novel effects.
\\\par
Klingbeil~\cite{klingbeil2005software} describes the techniques used in a high quality graphical
spectral editing and resynthesis program called SPEAR\@. The FFT analysis is done taking into
account the fundamental frequency of the sound to analyze, so that the transform can be optimized
to accurately capture as many partials as possible. Spectral peaks are then selected and very quiet
peaks are discarded. A linear prediction method is used to follow the peaks over time. The data
structure for storage of spectral data involves division of time into a number of frames which keep
track of which partials are active during that time, so that the search for a specific partial can
be narrowed. User interface design choices are also covered in detail.

Although the interface and granularity of SPEAR differs significantly from those in ScPy, the
techniques for high quality resynthesis presented should prove useful for improving the quality of
the sound output.
\\\par
While the Fourier transform is by far the most commonly used technique for spectral manipulation in
digital audio effects, there exist numerous other transformations which may be useful as well to
achieve certain audio effects. One such technique is the Mellin transform, described in a audio
effects context by De Sena and Rocchesso~\cite{de2004fast}. The Mellin transform has a scale
invariance property that is analogous to the shift invariance property of the Fourier transform. It
can be computed quickly by applying two extra steps before doing a Fourier transform: exponential
time warping and exponential multiplication. A few simple audio effects are presented using the
Mellin transform, time stretching, filtering and random phase deviation (whisperization).

As the Mellin transform can be implemented simply as an extension to the Fourier transform, it
should be fairly easy to add support for it in ScPy, and it will hopefully enable the use of
some different audio effects than can be achieved easily with the Fourier transform. However, the
paper mentions that a short-time Mellin transform has not been attempted, so some further work may
need to be done in that area first.
\\\par
Collins et al.~\cite{collins2003live} present an introduction to the field of live coding, where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the just in time library.

@@

\subsection{Lanugages}

@@

The ideas in Nilson's paper shed light on some of the challenges faced in this project.
One such challenge is the amount of writing required to produce an interesting result from a blank
slate. We intend to minimise this by providing many effect primitives and composition
mechanisms. However, it will be important to ensure the language is accessible to ensure it can be
quickly learned to a level of expertise where lack of technical knowledge is not an impediment.
\\\par
Domain-Specific Languages (DSLs) are small programming languages designed for use in a specific
problem domain. They often lack general purpose programming constructs and have a syntax which is
very concise for describing programs in their domain. Van Deursen et al.~\cite{van2000domain}
present a survey of available DSL literature, along with the pros and cons of using a DSL, example
DSLs, design methodology and implementation approaches. The biggest advantage of DSLs is that they
enable expression of solutions in a way that is common for the problem domain. However, they can be
costly to develop and difficult to design. Implementation techniques include the classical approach
of interpretation or compilation, embedded languages / domain-specific libraries, preprocessing, or
compiler extension. Additionally, the references of the paper each include a short summary.

This review reveals many of the possible options for the DSL used within the project. Although we
decided not to design a DSL for our project, analyzing the design and implementation techniques
presented here will be helpful for future work in this area.
\\\par
As Python is an easily extendible general purpose language, it has applications in many scientific
fields. Glover et al.~\cite{glover2011python} present a discussion of the use of Python for audio
signal processing. They note that while audio processing libraries exist for many languages,
scripting languages such as Python or MATLAB see widespread adoption because of the speed at which
prototypes can be built. General purpose Python libraries such as NumPy, SciPy and Matplotlib
include many essential functions for audio processing, such as file IO, Fourier transforms, signal
processing and interpolation. There are additionally more specific libraries such as Simpl, a
sinusoidal modelling library, and Modal, an onset detection library. Any of these libraries can
easily be used within ScPy simply by installing and importing them.
\\\par
Another open-source Python library which could be used in ScPy is librosa~\cite{mcfee2015librosa}
by McFee et al.\ librosa is a music signal analysis library designed to help transition music
information retrieval (MIR) researchers into using Python, as well as make MIR techniques readily
available to all Python programmers.\ librosa's core design choices and conventions are documented
in detail to inspire future library authors to follow similarly good coding practice. As well as
standard audio analysis functions, librosa also includes support for effects and audio IO.\@

librosa has quite an extensive list of features and would likely cover all of the audio related
functionality which is missing from NumPy. It can be used with ScPy simply by installing and importing
it.
\\\par
The ixi lang~\cite{magnusson2011ixi} is a live coding language. It is implemented as an interpreter
built in SuperCollider, a choice which allows for SuperCollider and ixi lang code to be mixed
together in a similar manner to ScPy. The ixi lang is built to reduce the startup time in a live
coding performance and enable faster musical experimentation through syntax more concise than
SuperCollider. The paper also contains user feedback on the usability of the ixi lang.

@@

The choice of SuperCollider was largely because it is one of the most widely used audio programming
environments. Using SuperCollider also enabled the reuse of many parts of its code base. Since it
already has STFT based operations, it was a good target for extension purposes in that area.
